---
title: "WICHE R Package"
author: "Matt Campbell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`wicher` is a package used to make working with the WICHE enrollment
projections easier in R. We generally use WICHE enrollments to figure
out the percentage of students who participate in our assessments.

We also use WICHE to visualize demographics in our Excel dashboards so I'm
putting some examples of visualizing in R below. I think this might make it
easier to prototype the graphics before trying to work them into the dashboard.

### Installing WICHE R Package

You can install `wicher` from my github repository like
this:

``` r

# MAKE SURE YOU HAVE DEVTOOLS INSTALLED
# AND REFERENCED

install.packages("devtools")
library(devtools)

# USE DEVTOOLS TO INSTALL THE WICHE PACKAGE 

install_github('mattjcamp/wicher','mattjcamp')
library(wicher)

# REFERENCE GGPLOT2

install.packages("ggplot2")
library(ggplot2)

```

You will need to install the packages above if you want to reproduce 
these examples.

### Datasets Included in WICHE R Package

`wiche_graduate_projections` is the built in dataframe that comes with
this package. It looks like this:

```{r warning = FALSE, message = FALSE, echo = TRUE}

library(wicher)
library(tidyverse)

head(wiche_graduate_projections)

```

Usually I use this dataframe to `JOIN` to assessment 
data aggregated by state and year to get participation rates, but sometimes 
you may want to just lookup a few datapoints.

### Simple Visualizations

So I just started adding visualizations using R since we 
mostly use Excel to produce graphics. Still, it's nice to 
visualize data right in R while we are doing analysis. When I'm
not too worried about the appearence of the graphics I just use
`qplot` (as in *quick* plot). This is part of `ggplot2` that 
we added at the top of this doc.

#### Showing Trends

Here is how to do a quick line plot to show trends. So I pulled the
enrollments for White and Hispanic students in California for the
years 2000 to 2010. Then I used `qplot` to make a chart that would 
show the trend in enrollments for these two groups.

```{r fig.width = 7, fig.height = 4, echo = TRUE}

library(ggplot2)

enrollments <- 
  wiche_graduate_projections %>% 
  filter(year %in% 2000:2010,
         location %in% "ca",
         race %in% c("white", "hispanic"),
         grade == "g") %>% 
  arrange(year, race)

qplot(year, n, data = enrollments,
      color = race,
      geom = c("point", "smooth"),
      main = "California Enrollment",
      xlab = "Academic Year",
      ylab = "Enrollment")

```      

>**NOTE** I found that these plots look terrible unless you set the right
figure dimensions in the markdown code. Look at the code file to see how this is
done.

Here is the same thing but using the full `ggplot` function:

```{r fig.width = 7, fig.height = 4, echo = TRUE}

ggplot(enrollments, aes(x = as.character(year), y = n, group = race)) +
  geom_smooth(aes(color = race)) +
  geom_point(aes(color = race)) +
  ggtitle("California Enrollment") +
  labs(x = "Academic Year", y = "Enrollment")

```      

>**NOTE** I'm putting this here as a reference for when we want something
like this and we want to control the output. `ggplot2` has more options that `qplot`.

#### Comparing Groups

Here is an example of using column charts to compare groups. We will look again
at California at the beginning and the end of the trend examined above.

First, make the chart for 2005:

```{r fig.width = 7, fig.height = 4, echo = TRUE}

enrollments <- 
  wiche_graduate_projections %>% 
  filter(year %in% 2005,
         location %in% "ca",
         race %in% c("white", "asian", "hispanic", "native", "black"),
         grade == "g") %>% 
  arrange(year, race)

ggplot(data = enrollments, aes(race, n)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("California Enrollment (2005)") +
  labs(x = "Academic Year", y = "Enrollment")

```
