---
title: "WICHE R Package"
author: "Matt Campbell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`wicher` is a package used to make working with the WICHE enrollment
projections easier in R. We generally use WICHE enrollments to figure
out the percentage of students who participate in our assessments.

We also use WICHE to visualize demographics in our Excel dashboards so I'm
putting some examples of visualizing in R below. I think this might make it
easier to prototype the graphics before trying to work them into the dashboard.

### Installing WICHE R Package

You can install `wicher` from my github repository like
this:

``` r

# MAKE SURE YOU HAVE DEVTOOLS INSTALLED
# AND REFERENCED

install.packages("devtools")
library(devtools)

# USE DEVTOOLS TO INSTALL THE WICHE PACKAGE ALONG
# WITH CODERR (WICHE USES THIS PACKAGE)

install_github('mattjcamp/coderr','mattjcamp')
install_github('mattjcamp/wicher','mattjcamp')

# REFERENCE GGPLOT2

install.packages("ggplot2")
library(ggplot2)

# REFERENCE WICHE PACKAGE FOR REST OF ANALYSIS

library(wicher)

```

You will need to install the packages above if you want to reproduce 
these examples.

### Datasets Included in WICHE R Package

Today, there is only one dataset in this package: `wiche_2013`. This is the
source that I've used for a bunch of cohort analysis over the past two years.

Here is what the table looks like:

```{r warning = FALSE, message = FALSE, echo = TRUE}

library(wicher)

head(wiche_2013)

```

>**NOTE** As soon as the new projections are available, we will be making a new 
dataset for 2016.

Usually I use this dataframe to `JOIN` to assessment 
data aggregated by state and year to get participation rates, but sometimes 
you may want to just lookup a few datapoints.

### Enrollment Lookups

Use the function `lookup_wiche` to look up specific enrollment information.
You can specify filters for `location`, `sector`, `race`, `gender`, `grade` and
`year`. For example, if you want to see enrollments for 2012 to 2016 for males
and females you would do this:

```{r warning = FALSE, message = FALSE, echo = TRUE}

lookup_wiche(year = 2012:2016,
             gender = c("f", "m"))

```

The paramaters all have defaults and since you didn't specify
`location` the output defaulted to **us** for United States.
To see the same thing but for Pennsylvania and New York you 
would do this:

```{r warning = FALSE, message = FALSE, echo = TRUE}

lookup_wiche(location = c("pa","ny"),
             year = 2012:2016,
             gender = c("f", "m"))

```

### Simple Visualizations

So I just started adding visualizations using R since we 
mostly use Excel to produce graphics. Still, it's nice to 
visualize data right in R while we are doing analysis. When I'm
not too worried about the appearence of the graphics I just use
`qplot` (as in *quick* plot). This is part of `ggplot2` that 
we added at the top of this doc.

#### Showing Trends

Here is how to do a quick line plot to show trends. So I pulled the
enrollments for White and Hispanic students in California for the
years 2000 to 2010. Then I used `qplot` to make a chart that would 
show the trend in enrollments for these two groups.

```{r fig.width = 7, fig.height = 4, echo = TRUE}

library(ggplot2)

enrollments <- lookup_wiche(year = 2000:2010,
                            location = "ca",
                            race = c("white", "hispanic"))

qplot(year, n, data = enrollments, 
      color = race, 
      geom = c("point", "smooth"),
      main = "California Enrollment",
      xlab = "Academic Year", 
      ylab = "Enrollment")

```      

I choose these data points because I knew from our previous analysis that they
would show an interesting trend.

>**NOTE** I found that these plots look terrible unless you set the right
figure dimensions in the markdown code. Look at the code file to see how this is
done.

Here is the same thing but using the full `ggplot` function:

```{r fig.width = 7, fig.height = 4, echo = TRUE}

ggplot(enrollments, aes(x = as.character(year), y = n, group = race)) +
  geom_smooth(aes(color = race)) +
  geom_point(aes(color = race)) + 
  ggtitle("California Enrollment") +
  labs(x = "Academic Year", y = "Enrollment")

```      

>**NOTE** I'm putting this here as a reference for when we want something
like this and we want to control the output.

#### Comparing Groups

Here is an example of using column charts to compare groups. We will look again
at California at the beginning and the end of the trend examined above.

First, make the chart for 2000:

```{r fig.width = 7, fig.height = 4, echo = TRUE}

enrollments <- lookup_wiche(year = 2000,
                            location = "ca",
                            race = c("white", "asian", "hispanic", "native", "black"))

ggplot(data = enrollments, aes(race, n)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("California Enrollment (2000)") +
  labs(x = "Academic Year", y = "Enrollment")

```

Now, do the same thing for 2010:

```{r fig.width = 7, fig.height = 4, echo = TRUE}

enrollments <- lookup_wiche(year = 2010,
                            location = "ca",
                            race = c("white", "asian", "hispanic", "native", "black"))

ggplot(data = enrollments, aes(race, n)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("California Enrollment (2010)") +
  labs(x = "Academic Year", y = "Enrollment")

```

## Next Steps

- Add 2016 enrollment projections to package
- Add geographic visualization templates
- Prototype new visualizations for the dashboard
- Add 2008 projections since some of the research I've seeing is referencing projections back to 1992
